---
title: "Genome-wide association study for quantitative traits"
author: "Elise Tourrette, Timothée Flutre, Bertrand Servin"
date: "March 12, 2025"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
editor: visual
execute:
  freeze: auto
  cache: true
---

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(ggbeeswarm)
library(broom)

library(qvalue)
library(ashr)
library(rrBLUP)
library(gaston)
```

# Main objective

Studying the **genetic architecture** of complex (quantitative) traits aims at **idenfitying polymorphisms** (number, genomic locations, effects) affecting the phenotypic differences between individuals.

This can help to:

-   understand the biological processes underlying phenotypes
-   predict individual genetic values (diagnostic, selection ...)
-   separate out genetics from environmental effects

Before the 2000's, this was done thourgh **linkage analysis**, studying segregation of alleles in families. This approach although robust, was clearly limited in particular due to its lack of power, see [Risch and Merichangas (1996)](https://science.sciencemag.org/content/273/5281/1516).

A more powerful approach consists in sampling *unrelated* individuals from a general population and test for **association** between individual genotypes and their phenotypes = GWAS.

```{r message=FALSE}
## Sample Size
N <- 200
## QTL parameters: allele frequency and effect
f <- 0.2
beta <- 1
## sample genotypes
G <- rbinom(N, 2, f)
Y <- rnorm(N, mean = G * beta)
ggplot(tibble(G = G, Y = Y), aes(x = G, y = Y)) +
  geom_beeswarm() +
  theme_minimal() +
  geom_smooth(method = "lm") +
  xlab("Genotype") +
  ylab("Phenotype")
```

The effect of the marker on the phenotype can be estimated by a simple linear model

```{r}
mod <- lm(Y ~ G)
summary(mod)
```

Although this appears simple enough, devil is in the details, and today we will be looking into how to deal with them in order to perform a robust yet powerful GWAS.

# Testing for association with a single QTL

## Linkage Disequilibrium and Marker Density

Because we work with samples of unrelated individuals, the ability to detect QTLs relies on the **Linkage Disequilibrium** between genotyped loci (markers) and acutal QTLs. Indeed, if the genotype correlation between marker and QTL is *r*, the power to detect the QTL with the marker decreases with $r^2$. So one of the most important factor affecting the power of a GWAS is the **marker density**.

We will illustrate this by performing a simple GWAS on two datasets of different density

### low density GWAS

```{bash, eval = FALSE}
## suppose that we are in the student's working directory
mkdir -p simulations
cd simulations

## Simulate a dataset with 200 markers per chromosome, one of which is a QTL
python ../QLife2025/simulations/src/main.py -savedFolder 'gwas_simu1' -nQTL 1 -Lchr 200 -LG 100 -varEffect 1.0 -proportion0 0.0 -varEffect0 0.0 -optim 0 -varW 100 -h2 0.5 -G 1 -N 100 -Npop 10000 -nTrait 1 -nChr 1 -mu 1e-5 >> out_gwas &  
```

```{r}
## test for association between markers and QTLs with a simple linear regression
## use broom ?

# setwd("simulations/gwas_simu1")

# load the data
# pedigree information, also contains the phenotypes and breeding values
ped <- read_table("simulations/gwas_simu1/pedigree.txt")

# genotypes
geno <- read.table("simulations/gwas_simu1/genotype.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE, row.names = 1, check.names = FALSE)

## only look at the initial generation
indKeep <- which(ped$generation == 0)
ped <- ped[indKeep, ]
geno <- geno[indKeep, ]

# snp information
snp <- read_table("simulations/gwas_simu1/SNP_INFO.txt")
nQTL <- nrow(snp |> filter(beta_trait_1 != 0)) # number of QTLs
nSNP <- nrow(snp) # total number of SNPs

# frequency of allele 1 (allele with effect)
freq <- colMeans(geno) / 2
# filter the SNPs
geno <- geno[, freq > 0 & freq < 1]
snp <- snp[freq > 0 & freq < 1, ]

# association: linear regression, independently for each markers
lm.res <- apply(geno, 2, function(x) summary(lm(ped$pheno1 ~ x))$coeff[2, ])
# reformat the result for easier plotting
lm.res <- as.data.frame(t(lm.res))
colnames(lm.res) <- c("beta_hat", "se", "tstat", "pval")
lm.res$beta <- snp$beta_trait_1
lm.res$snp_id <- snp$snp_id
lm.res$qtl <- 0
lm.res$qtl[lm.res$beta != 0] <- 1
nsnp <- nrow(lm.res)

# ggplot(lm.res) +
#    geom_point(aes(x = beta, y = beta_hat, col = as.factor(qtl))) +
#    scale_colour_manual(values = c('0' = "black", '1' = "red"))

ggplot(lm.res) +
  geom_point(aes(x = snp_id, y = -log10(pval), col = as.factor(qtl))) +
  scale_colour_manual(values = c("0" = "black", "1" = "red"))

# distribution of p-values
# hist(lm.res$pval, breaks = 100, prob = TRUE)
```

**If you had not included the QTL in your analysis, would you have found it ?**

### High density GWAS

```{bash, eval = FALSE}
## Simulate a dataset with 20000 markers per chromosome, one of which is a QTL
python ../QLife2025/simulations/src/main.py -savedFolder 'gwas_simu2' -nQTL 1 -Lchr 20000 -LG 10 -varEffect 1.0 -proportion0 0.0 -varEffect0 0.0 -optim 0 -varW 100 -h2 0.5 -G 1 -N 100 -Npop 10000 -nTrait 1 -nChr 1 -mu 1e-5 >> out_gwas &
```


```{r}
## test for association between markers and QTLs with a simple linear regression

# setwd("simulations/gwas_simu2")

# load the data
# pedigree information, also contains the phenotypes and breeding values
ped <- read.table("simulations/gwas_simu2/pedigree.txt", header = TRUE)
ped$generation <- factor(as.character(ped$generation), levels = as.character(sort(unique(ped$generation))), ordered = TRUE)
# genotypes
geno <- read.table("simulations/gwas_simu2/genotype.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE, row.names = 1, check.names = FALSE)

## only look at the initial generation
indKeep <- which(ped$generation == 0)
ped <- ped[indKeep, ]
geno <- geno[indKeep, ]

# snp information
snp <- read.table("simulations/gwas_simu2/SNP_INFO.txt", header = TRUE, sep = "\t")
nQTL <- nrow(snp[snp$beta_trait_1 != 0, ]) # number of QTLs
nSNP <- nrow(snp) # total number of SNPs

# frequency of allele 1 (allele with effect)
freq <- colMeans(geno) / 2
# filter the SNPs
geno <- geno[, freq > 0 & freq < 1]
snp <- snp[freq > 0 & freq < 1, ]

# association: linear regression, independently for each markers
lm.res <- apply(geno, 2, function(x) summary(lm(ped$pheno1 ~ x))$coeff[2, ])
# reformat the result for easier plotting
lm.res <- as.data.frame(t(lm.res))
colnames(lm.res) <- c("beta_hat", "se", "tstat", "pval")
lm.res$beta <- snp$beta_trait_1
lm.res$snp_id <- snp$snp_id
lm.res$qtl <- 0
lm.res$qtl[lm.res$beta != 0] <- 1
nsnp <- nrow(lm.res)

# ggplot(lm.res) +
#  geom_point(aes(x = beta, y = beta_hat, col = as.factor(qtl))) +
#  scale_colour_manual(values = c("0" = "black", "1" = "red"))

ggplot(lm.res) +
  geom_point(aes(x = snp_id, y = -log10(pval), col = as.factor(qtl))) +
  scale_colour_manual(values = c("0" = "black", "1" = "red"))

# distribution of p-values
# hist(lm.res$pval, breaks = 100, prob = TRUE)
```

**If you had not included the QTL in your analysis, would you have found it ?**

Generally, your study system will come with its available genotyping tools. These will constrain the marker density of your GWAS. In addition to genotyping and sequencing techniques (RADseq, Low pass sequencing ...), statistical methods can be used to improve marker density: **genotype imputation methods**. We will not cover them in this practical, but know that if you have sufficiently large diversity panels for your species, these methods can greatly increase power of GWAS by predicting the genotype of individuals for large number of known markers.

## Multiple testing: Family Wise Error Rate

By increasing the number of markers assessed for association, we increase the number of statistical tests. In order to call significant associations we need to account for this multiple testing.

The simplest way to control for multiple testing is to control the Family Wise Error Rate (FWER) by the Bonferroni procedure. The FWER is the probability that one of the test deemed significant is a false positive. For a single test, this is simply the associated p-value ($p$). So if you want to control the FWER at level $\alpha$, you will call your single test significant if $p \leq \alpha$.

**Add a note on the null distribution of p-values** However, if you perform $n$ independant statistical test, you want to control the FWER with the same level $\alpha$, you need to adjust your threshold for each of the $n$ tests. One procedure to do this is the so-called Bonferroni procedure, which corresponds to setting the significance threshold at $\alpha/n$. So for example if you want to control FWER at $\alpha=0.05$ and perform 1 million tests for association, you should only reject tests that have a pvalue less than $0.05 \times 10^{-6} = 5.10^{-8}$

Implement the Bonferroni procedure on your GWAS results, how many SNPs can you call significant ?

```{r}
## Implement the Bonferroni procedure on the GWAS results
alpha <- 0.05
ggplot(lm.res) +
  geom_point(aes(x = snp_id, y = -log10(pval), col = as.factor(qtl))) +
  geom_hline(yintercept = -log10(alpha / nsnp), col = "green") + # threshold corrected for the number of tests (i.e. number of SNPs tested)
  scale_colour_manual(values = c("0" = "black", "1" = "red"))
```

One problem with the Bonferroni procedure is that it is **conservative**, in particular in the presence of correlated tests. One way to avoid this is to perform random **permutations** of the genotype / phenotype associations. This effectively simulates data under the null (no association between phenotype and genotype) and allows to account both for multiple testing and correlation between tests.

```{r, eval=FALSE}
## assess significance via permutations

# permutations:
# permutate the phenotypes
# and test the association as before
# for each SNP, look at the number of times the observed p-value is LESS than the minimal p-value obtained for a permutation (all test - i.e. SNP - considered)
# do it multiple (nrep) times

permutation_test <- function(ped, geno, nrep, pval_obs) {
  res <- rep(1, length(pval_obs))
  for (irep in 1:nrep) {
    print(irep)
    pheno <- sample(ped$pheno1)
    pm.res <- apply(geno, 2, function(x) summary(lm(pheno ~ x))$coeff[2, ])
    pm.res <- as.data.frame(t(pm.res))
    colnames(pm.res) <- c("beta_hat", "se", "tstat", "pval")
    res <- res + (pval_obs > min(pm.res$pval))
  }
  res <- res / (nrep + 1)
  return(res)
}

lm.res$pval2 <- permutation_test(ped, geno, 100, pval_obs = lm.res$pval)
```

```{r, eval =FALSE}
##
alpha <- 0.05
ggplot(lm.res) +
  geom_point(aes(x = snp_id, y = -log10(pval2), col = as.factor(qtl))) +
  geom_hline(yintercept = -log10(alpha), col = "green") + # threshold corrected for the number of tests (i.e. number of SNPs tested)
  scale_colour_manual(values = c("0" = "black", "1" = "red"))
```

As you have seen, this procedure is quite time consuming (and we only
did 100 replicates, which we should increase at least by 10). Also it only works 
because individuals are *exchangeable* as they are unrelated. In large sample
sizes, this exchangeability hypothesis is not true anymore (kinship varies
between pairs of individuals) and the permutation scheme cannot be used.

## Population structure

Until now, we have assumed (and simulated) individuals that come from a random mating population of constant size with no selection or migration. In this ideal situation all individuals are essentially exchangeable and we have no factors counfounding the association between genotype and phenotype.

Unfortunately, almost no real world scenario fits this situation. So one has to account for **population structure** when performing a GWAS. There are several ways to do it, we will see one in this section and another one in the next.

First, let's run some new simulations where our initial panmictic populations is subjected to selection (**ADD DETAILS ON THE SIMULATIONS HERE**) and see the consequences on our GWAS results.

### Simulation 

```{bash, eval = FALSE}
## RUN simulations with selection
python ../QLife2025/simulations/src/main.py -savedFolder 'gwas_simu3' -nQTL 1 -Lchr 2000 -LG 100 -varEffect 1.0 -proportion0 0.0 -varEffect0 0.0 -optim 1 -varW 0.01 -h2 0.5 -G 10 -N 100 -Npop 10000 -nTrait 1 -nChr 10 -mu 1e-5 >> out_gwas &
```

```{r}
## Run the lm GWAS
# setwd("simulations/gwas_simu3")

# load the data
# pedigree information, also contains the phenotypes and breeding values
ped <- read.table("simulations/gwas_simu3/pedigree.txt", header = TRUE)
ped$generation <- factor(as.character(ped$generation), levels = as.character(sort(unique(ped$generation))), ordered = TRUE)
# genotypes
geno <- read.table("simulations/gwas_simu3/genotype.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE, row.names = 1, check.names = FALSE)

## only look at the initial generation
# indKeep <- which(ped$generation > 0)
# ped <- ped[indKeep, ]
# geno <- geno[indKeep, ]

# snp information
snp <- read.table("simulations/gwas_simu3/SNP_INFO.txt", header = TRUE, sep = "\t")
print(nrow(snp[snp$beta_trait_1 != 0, ])) # number of QTLs
print(nrow(snp)) # total number of SNPs

# frequency of allele 1 (allele with effect)
freq <- colMeans(geno) / 2
# filter the SNPs
geno <- geno[, freq > 0 & freq < 1]
snp <- snp[freq > 0 & freq < 1, ]

# association: linear regression, independently for each markers
lm.res <- apply(geno, 2, function(x) summary(lm(ped$pheno1 ~ x))$coeff[2, ])
# reformat the result for easier plotting
lm.res <- as.data.frame(t(lm.res))
colnames(lm.res) <- c("beta_hat", "se", "tstat", "pval")
lm.res$beta <- snp$beta_trait_1
lm.res$snp_id <- snp$snp_id
lm.res$chr_id <- snp$chr_id
lm.res$qtl <- 0
lm.res$qtl[lm.res$beta != 0] <- 1
nsnp <- nrow(lm.res)

## Look at the results
## ggplot(lm.res) +
##  geom_point(aes(x = beta, y = beta_hat, col = as.factor(qtl))) +
##  scale_colour_manual(values = c("0" = "black", "1" = "red"))

ggplot(lm.res) +
  geom_point(aes(x = snp_id, y = -log10(pval), shape = as.factor(qtl), col = as.factor(chr_id))) +
  geom_hline(yintercept = -log10(alpha / nsnp), col = "green")  # threshold corrected for the number of tests (i.e. number of SNPs tested)

# distribution of p-values
hist(lm.res$pval, prob = TRUE, n = 100)
```

-   Run the GWAS (using lm) on individuals as before
-   Using the Bonferroni procedure, how many SNPs would be called significant ?
-   Plot the distribution of p-values ? Compare it to the distribution of p-values in the GWAS in the random mating population. What do you see ?

**Population structure** can be caused by many processes, including selection, isolation-by-distance, barriers to gene flow, assortative mating ... It will create false positives in GWAS if not accounted for properly.

To account for populations structure, the simplest method is to use Principal Component Analysis of the genotype matrix to reveal axes of genetic structuration. We can then include the main PCs as regressors in the GWAS model

```{r}
# # do the PCA on the GRM (cf. Genomic Prediction practical)
# # use the Van Raden estimation
# estimVanRanden1 <- function(mat, AFs = NULL) {
#   if (all(mat >= 0)) { # need column-wise mean-centering
#     M <- mat
#     if (is.null(AFs)) {
#       AFs <- colMeans(mat) / 2
#     } else {
#       stopifnot(length(AFs) == ncol(mat))
#     }
#     tmp <- matrix(rep(1, nrow(M))) %*% (2 * AFs)
#     Z <- M - tmp
#   } else { # columns assumed already centered
#     stopifnot(sapply(colMeans(mat), all.equal, 0.0))
#     Z <- mat
#   }
#   tcrossprod(Z, Z) / (2 * sum(AFs * (1 - AFs)))
# }
#
# AFs_VR1 <- colMeans(geno) / 2 ## get the allele frequencies
# tmp <- matrix(rep(1, nrow(geno))) %*% (2 * AFs_VR1)
# Z_VR1 <- as.matrix(geno - tmp)
# GRM_VR1 <- estimVanRanden1(Z_VR1, AFs_VR1)

## Perform a PCA on the genotype matrix
# pca <- eigen(GRM_VR1, symmetric = TRUE)

pca <- svd(geno)

## looking at Eigen Values, decide on a number of component to include
plot(100 * pca$d / sum(pca$d), xlab = "PC", ylab = "% Variance explained (GRM)", xlim = c(1, 20), type = "h")

## add these components in the lm
nPC <- 20 # number of PC to use
# association testing
lm.res.pca <- apply(geno, 2, function(x) summary(lm(ped$pheno1 ~ pca$u[, 1:nPC] + x))$coeff[nPC + 2, ])
# reformat the result
lm.res.pca <- as.data.frame(t(lm.res.pca))
colnames(lm.res.pca) <- c("beta_hat", "se", "tstat", "pval")
lm.res.pca$beta <- snp$beta_trait_1
lm.res.pca$snp_id <- snp$snp_id
lm.res.pca$chr_id <- snp$chr_id
lm.res.pca$qtl <- 0
lm.res.pca$qtl[lm.res.pca$beta != 0] <- 1
nsnp <- nrow(lm.res.pca)

ggplot(lm.res.pca) +
  geom_point(aes(x = snp_id, y = -log10(pval), shape = as.factor(qtl), col = as.factor(chr_id))) +
  geom_hline(yintercept = -log10(0.05 / nsnp), col = "green")  ## threshold corrected for the number of tests (i.e. number of SNPs tested)


# distribution of p-values
hist(lm.res.pca$pval, prob = TRUE, n = 100)
```

-   Run the GWAS (using lm + PC covariates) on individuals from the last generation of the dataset
-   Using the Bonferroni procedure, how many SNPs would be called significant ?
-   Plot the distribution of p-values ? Compare it to the distribution of p-values in the GWAS without PC correction. What happened ?

# Multiple QTLs

Up until now, we have considered a single QTL participating in the phenotype.
We have tested markers one at a time to test for association. In a more general
setting, in particular in the case of **polygenic** adaptation, we want to 
consider situations where multiple QTLs contribute to the trait. In the second
part of the workshop we will be simulating more complex genetic architecture
and use different approaches to infer them. 

## Oligogenic model

First, we will be considering a situation were there are multiple but still
relatively few QTLs segregating in the population. This corresponds to an 
*oligogenic* determinism. The questions we want to answer are how many QTLs
are there, what are their global effect on the trait of interest.

### Simulations

First, we will simulate a larger (n=1000) sample from a panmictic population of effective size 
$Ne = 10 000$. We will simulate arount 10,000 polymorphisms out of which only
about 100 are going to be QTLs. The overall heritability of the trait is set to 
50%. 

```{bash, eval = FALSE}
## Simulations for the multi-QTLs
# QTLs + neutral SNPs
python ../QLife2025/simulations/src/main.py -savedFolder 'gwas_simu4' -nQTL 100 -Lchr 2000 -LG 100 -varEffect 100.0 -proportion0 0.0 -varEffect0 0.0 -optim 0 -varW 100 -h2 0.5 -G 1 -N 1000 -Npop 10000 -nTrait 1 -nChr 10 -mu 1e-5 >> out_gwas &

```

As before, we read in the data from the simulation results.

```{r}
## load the data
## setwd("simulations/gwas_simu4")

# load the data
# pedigree information, also contains the phenotypes and breeding values
ped <- read.table("simulations/gwas_simu4/pedigree.txt", header = TRUE)
ped$generation <- factor(as.character(ped$generation), levels = as.character(sort(unique(ped$generation))), ordered = TRUE)
# genotypes
geno <- read.table("simulations/gwas_simu4/genotype.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE, row.names = 1, check.names = FALSE)

## only look at the initial generation
# indKeep <- which(ped$generation == 0)
# ped <- ped[indKeep, ]
# geno <- geno[indKeep, ]

# snp information
snp <- read.table("simulations/gwas_simu4/SNP_INFO.txt", header = TRUE, sep = "\t")
print(paste("nQTL =", nrow(snp[snp$beta_trait_1 != 0, ]))) # number of QTLs
print(paste("nSNP =", nrow(snp))) # total number of SNPs

# frequency of allele 1 (allele with effect)
freq <- colMeans(geno) / 2
# filter the SNPs
geno <- geno[, freq > 0 & freq < 1]
snp <- snp[freq > 0 & freq < 1, ]
```

As the population is panmictic, we can expect individuals to be approximately
unrelated and work with a classical linear regression approach for the GWAS.

```{r}
## linear regression

# association: linear regression, independently for each markers
lm.res <- apply(geno, 2, function(x) summary(lm(ped$pheno1 ~ x))$coeff[2, ])
# reformat the result for easier plotting
lm.res <- as.data.frame(t(lm.res))
colnames(lm.res) <- c("beta_hat", "se", "tstat", "pval")
lm.res$beta <- snp$beta_trait_1
lm.res$snp_id <- snp$snp_id
lm.res$chr_id <- snp$chr_id
lm.res$qtl <- 0
lm.res$qtl[lm.res$beta != 0] <- 1
nsnp <- nrow(lm.res)

ggplot(lm.res) +
  geom_point(aes(x = beta, y = beta_hat, col = as.factor(qtl))) +
  scale_colour_manual(values = c("0" = "black", "1" = "red"))

ggplot(lm.res) +
  geom_point(aes(x = snp_id, y = -log10(pval), shape = as.factor(qtl), col = as.factor(chr_id))) 

# distribution of p-values
hist(lm.res$pval, breaks = 100, prob = TRUE)

## qqplot
qqnorm(lm.res$tstat) ## t-stat
qqline(lm.res$tstat, col = "red")
## qqplot for -log10(pval)
qqplot(-log10(ppoints(nsnp)), -log10(lm.res$pval), xlab = "theoretical", ylab = "observed", main = "Q-Q Plot for -log10 Pval")
abline(a = 0, b = 1, col = "red")
```
### Multiple Testing: False Discovery Rate 
q-values

```{r}
qval <- qvalue(p = lm.res$pval)
summary(qval) ## pi0: overall proportion of true null hypothesis (beta = 0)
plot(qval)
# plot(-log10(qval$pvalues), -log10(qval$qvalues))
#
# lm.res$qval <- qval$qvalues
#
# ggplot(lm.res) +
#   geom_point(aes(x = snp_id, y = -log10(qval), col = as.factor(qtl))) +
#   scale_colour_manual(values = c("0" = "black", "1" = "red"))
#
# ## distribution of q-values
# hist(lm.res$qval, prob = TRUE)
```

ashr

```{r}
beta.ash <- ash(lm.res$beta_hat, lm.res$se)
summary(beta.ash)

lm.res$beta_ash <- beta.ash$result$PosteriorMean
lm.res$qval_ash <- beta.ash$result$qvalue

ggplot(lm.res) +
  geom_point(aes(x = beta_hat, beta_ash, col = as.factor(qtl))) +
  scale_colour_manual(values = c("0" = "black", "1" = "red"))

ggplot(lm.res) +
  geom_point(aes(x = snp_id, y = -log10(qval_ash), shape = as.factor(qtl), col = as.factor(chr_id))) 
```

## Polygenic model

### Simulations

```{bash, eval=FALSE}
# some QTLs with large effects + QTLs with weak effects 
python ../QLife2025/simulations/src/main.py -savedFolder 'gwas_simu5' -nQTL 100 -Lchr 2000 -LG 100 -varEffect 100.0 -proportion0 1.0 -varEffect0 0.01 -optim 0.2 -varW 1 -h2 0.5 -G 10 -N 100 -Npop 10000 -nTrait 1 -nChr 10 -mu 1e-5 >> out_gwas &
```

Read in data:

```{r}
## load the data
## setwd("simulations/gwas_simu4")

# load the data
# pedigree information, also contains the phenotypes and breeding values
ped <- read.table("simulations/gwas_simu5/pedigree.txt", header = TRUE)
ped$generation <- factor(as.character(ped$generation), levels = as.character(sort(unique(ped$generation))), ordered = TRUE)
# genotypes
geno <- read.table("simulations/gwas_simu5/genotype.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE, row.names = 1, check.names = FALSE)

## only look at the initial generation
indKeep <- which(ped$generation == 0)
ped <- ped[indKeep, ]
geno <- geno[indKeep, ]

# snp information
snp <- read.table("simulations/gwas_simu5/SNP_INFO.txt", header = TRUE, sep = "\t")
print(nrow(snp[snp$beta_trait_1 != 0, ])) # number of QTLs
print(nrow(snp)) # total number of SNPs

# frequency of allele 1 (allele with effect)
freq <- colMeans(geno) / 2
# filter the SNPs
geno <- geno[, freq > 0 & freq < 1]
snp <- snp[freq > 0 & freq < 1, ]
```

### Linear Regression

```{r}
# association: linear regression, independently for each markers
lm.res <- apply(geno, 2, function(x) summary(lm(ped$pheno1 ~ x))$coeff[2, ])
# reformat the result for easier plotting
lm.res <- as.data.frame(t(lm.res))
colnames(lm.res) <- c("beta_hat", "se", "tstat", "pval")
lm.res$beta <- snp$beta_trait_1
lm.res$snp_id <- snp$snp_id
lm.res$chr_id <- snp$chr_id
lm.res$qtl <- 0
lm.res$qtl[lm.res$beta != 0] <- 1
nsnp <- nrow(lm.res)
```

### Ridge Regression : Linear Mixed Effects (BLUP)

rrBLUP

```{r}
rrblup.res <- mixed.solve(y = ped$pheno1, Z = as.matrix(geno)) ## by default K = I
lm.res$beta_blup <- rrblup.res$u

ggplot(lm.res, aes(x = beta_hat, y = beta_blup)) +
  geom_point() +
  geom_abline()
```

### Linear Mixed Effects Model + QTLs

gaston

```{r}
## generate the inputs to be used
simu.fam <- cbind(fam.id = 1, ped[, c("ind_id", "father_id", "mother_id", "sex", "pheno1")])
colnames(simu.fam) <- c("famid", "id", "father", "mother", "sex", "pheno")
simu.bim <- cbind(snp[, c("chr_id", "snp_id", "gen_pos")], phys_pos = 10 * snp$gen_pos / snp$gen_pos[1], snp[, c("ALT", "REF")])
colnames(simu.bim) <- c("chr", "id", "dist", "pos", "A1", "A2")
simu.bim <- simu.bim[freq > 0 & freq < 1, ] ## remove the fixed SNPs
simu.gen <- as.matrix(geno)

x <- as.bed.matrix(simu.gen, simu.fam, simu.bim)
standardize(x) <- "p"
## calculate the GRM (needed for the background effect)
K <- GRM(x)
eigK <- eigen(K, symmetric = TRUE)
X <- matrix(1, nrow(x))

## association test
gaston.res <- association.test(x = x, Y = x@ped$pheno, X = X, method = "lmm", test = "wald", response = "quantitative", eigenK = eigK)

ggplot(gaston.res) +
  geom_point(aes(x = id, y = -log10(p), col = as.factor(chr_id)))

lm.res$beta_gaston <- gaston.res$beta
ggplot(lm.res) +
  geom_point(aes(x = beta_hat, y = beta_gaston))
```


### Bayesian Sparse Linear Mixed Effects
gemma (bslmm)

```{bash , eval=FALSE}
## reformat the input data
## to get the needed format

mkdir gemma
cd gemma

## as input, need .fam, .bim and .bed files
awk 'BEGIN {OFS="\t"} NR > 1 && $6 == 0 {print "1", $1, $3, $2, $4, $7}' ../pedigree.txt > simu.fam
LC_NUMERIC=C awk -F'\t' 'NR==2 {divisor = $3} NR > 1 {print $2, $1, $3, $3*10/divisor, $4, $5}' OFS='\t' ../SNP_INFO.txt > simu.bim
## there are 100 individuals in the founders, to adapt to the simulations
head -n 101 ../genotype.txt | awk 'NR > 1 { $1=""; print substr($0,2) }'  | tr -s ' ' '\t' | \
awk '
BEGIN { 
    FS = OFS = "\t" 
}
{
    for (i = 1; i <= NF; i++) {
        a[i] = a[i] $i "\t"
    }
}
END {
    for (i = 1; i in a; i++) {
        print substr(a[i], 1, length(a[i]) - 1)
    }
}
' > simu.txt 
## then use ldak (cf Genomic Prediction practical) to generate the .bed file
ldak6 --make-bed simu2 --gen simu.txt --bim simu.bim --fam simu.fam --gen-skip 0 --gen-headers 0 --gen-probs 1 --threshold 1

## run gemma (installed using conda)
gemma -bfile simu2 -bslmm 1

```


```{r}
## visualization

gemma.res <- read.table("simulations/gwas_simu5/gemma/output/result.param.txt", header = TRUE)
colnames(gemma.res) <- c("chr", "rs", "ps", "n_miss", "alpha_gemma", "beta_gemma", "gamma_gemma")

lm.res$rs <- as.numeric(row.names(lm.res))
data <- merge(lm.res, gemma.res)

ggplot(data, aes(x = beta_hat, y = alpha_gemma + beta_gemma)) +
  geom_point() +
  geom_abline()
```
