---
title: "Genome-wide association study for quantitative traits"
author: "Elise Tourrette, Timothée Flutre, Bertrand Servin"
date: "March 12, 2025"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: right
    number-sections: true
editor: visual
---

```{r echo=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggbeeswarm))
```

# Main objective

Studying the **genetic architecture** of complex (quantitative) traits aims at
**idenfitying polymorphisms** (number, genomic locations, effects) affecting the 
phenotypic differences between individuals. 

This can help to:

- understand the biological processes underlying phenotypes
- predict individual genetic values (diagnostic, selection ...)
- separate out genetics from environmental effects

Before the 2000's, this was done thourgh **linkage analysis**, studying segregation of alleles in families. This approach although robust,
was clearly limited in particular due to its lack of power, see [Risch and Merichangas (1996)](https://science.sciencemag.org/content/273/5281/1516). 

A more powerful approach consists in sampling *unrelated* individuals from a 
general population and test for **association** between individual genotypes and
their phenotypes = GWAS. 

```{r message=FALSE}
## Sample Size
N <- 200
## QTL parameters: allele frequency and effect
f <- 0.2
beta <- 1
## sample genotypes
G <- rbinom(N, 2, f)
Y <- rnorm(N, mean = G * beta)
ggplot(tibble(G = G, Y = Y), aes(x = G, y = Y)) +
  geom_beeswarm() +
  theme_minimal() +
  geom_smooth(method = "lm") +
  xlab("Genotype") +
  ylab("Phenotype")
```

The effect of the marker on the phenotype can be estimated by a simple linear model

```{r}
mod <- lm(Y ~ G)
summary(mod)
```

Although this appears simple enough, devil is in the details, and today we will
be looking into how to deal with them in order to perform a robust yet powerful
GWAS.

# Testing for association with a single QTL

## Linkage Disequilibrium and Marker Density

Because we work with samples of unrelated individuals, the ability to detect 
QTLs relies on the **Linkage Disequilibrium** between genotyped loci (markers) and
acutal QTLs. Indeed, if the genotype correlation between marker and QTL is *r*,
the power to detect the QTL with the marker decreases with **$r^2$**. 
So one of the most important factor affecting the power of a GWAS is the **marker density**. 

We will illustrate this by performing a simple GWAS on two datasets of different
density

### low density GWAS

```{r}
## Simulate a dataset with 200 markers per chromosome, one of which is a QTL

## test for association between markers and QTLs with a simple linear regression
## use broom ?
```

**If you had not included the QTL in your analysis, would you have found it ?** 

### High density GWAS

```{r}
## Simulate a dataset with 20000 markers per chromosome, one of which is a QTL

## test for association between markers and QTLs with a simple linear regression
```

**If you had not included the QTL in your analysis, would you have found it ?** 


Generally, your study system will come with its available genotyping tools. These 
will constrain the marker density of your GWAS. In addition to genotyping and
sequencing techniques (RADseq, Low pass sequencing ...), statistical methods
can be used to improve marker density: **genotype imputation methods**. We will
not cover them in this practical, but know that if you have sufficiently large
diversity panels for your species, these methods can greatly increase power
of GWAS by predicting the genotype of individuals for large number of known markers.

## Multiple testing

By increasing the number of markers assessed for association, we increase the 
number of statistical tests. In order to call significant associations we need 
to account for this multiple testing. 

The simplest way to control for multiple testing is to control the Family Wise 
Error Rate (FWER) by the Bonferroni procedure. The FWER is the probability that
one of the test deemed significant is a false positive. For a single test, this
is simply the associated p-value ($p$). So if you want to control the FWER at level
$\alpha$, you will call your single test significant if $p \leq \alpha$. 

**Add a note on the null distribution of p-values**
However, if you perform $n$ independant statistical test, you want to control
the FWER with the same level $\alpha$, you need to adjust your threshold for 
each of the $n$ tests. One procedure to do this is the so-called Bonferroni
procedure, which corresponds to setting the significance threshold at $\alpha/n$.
So for example if you want to control FWER at $\alpha=0.05$ and perform 1 million
tests for association, you should only reject tests that have a pvalue less 
than $0.05 \times 10^{-6} = 5.10^{-8}$

Implement the Bonferroni procedure on your GWAS results, how many SNPs can 
you call significant ?

```{r}
## Implement the Bonferronie procedure on the GWAS results
```

One problem with the Bonferroni procedure is that it is **conservative**, in 
particular in the presence of correlated tests. One way to avoid this is to
perform random **permutations** of the genotype / phenotype associations. This
effectively simulates data under the null (no association between phenotype and
genotype) and allows to account both for multiple testing and correlation between
tests. 

```{r}
## assess significance via permutations
```

## Population structure

Until now, we have assumed (and simulated) individuals that come from a random
mating population of constant size with no selection or migration. In this ideal
situation all individuals are essentially exchangeable and we have no factors 
counfounding the association between genotype and phenotype.

Unfortunately, almost no real world scenario fits this situation. So one has to
account for **population structure** when performing a GWAS. There are several
ways to do it, we will see one in this section and another one in the next. 

First, let's run some new simulations where our initial panmictic populations is
subjected to selection (**ADD DETAILS ON THE SIMULATIONS HERE**) and see the 
consequences on our GWAS results.

```{r}
## RUN simulations with selection

## Runt he lm GWAS

## Look at the results
```

- Run the GWAS (using lm) on individuals from the last generation of the dataset
- Using the Bonferroni procedure, how many SNPs would be called significant ?
- Plot the distribution of p-values ? Compare it to the distribution of p-values
in the GWAS in the random mating population. What do you see ? 

**Population structure** can be caused by many processes, including selection,
isolation-by-distance, barriers to gene flow, assortative mating ... It will
create false positives in GWAS if not accounted for properly.

To account for populations structure, the simplest method is to use Principal
Component Analysis of the genotype matrix to reveal axes of genetic structuration.
We can then include the main PCs as regressors in the GWAS model

```{r}
## Perform a PCA on the genotype matrix

## looking at Eigen Values, decide on a number of component to include

## add these components in the lm
```

- Run the GWAS (using lm + PC covariates) on individuals from the last 
generation of the dataset
- Using the Bonferroni procedure, how many SNPs would be called significant ?
- Plot the distribution of p-values ? Compare it to the distribution of p-values
in the GWAS without PC correction. What happened ? 

# Multiple QTLs
